---
title: "Analysis of NOAA Storm Data"
author: "Lalitha"
date: "December 28, 2017"
output: html_document
---

## Introduction

The purpose of this report is to explore the storm data collected by NOAA and analyse the natural disasters that cause most fatalities, injuries, property and crop damages.

## Data Processing
This step describes the process of downloading, cleaning and reading the storm data 

```{r}
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",temp)
data <- read.csv(bzfile(temp))
data$EVTYPE = toupper(data$EVTYPE)

reduced.data <- data[,c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "CROPDMG")]

head(data)

data[data$EVTYPE == "TSTM WIND", ]$EVTYPE = "THUNDERSTORM WIND"
data[data$EVTYPE == "THUNDERSTORM WINDS", ]$EVTYPE = "THUNDERSTORM WIND"
data[data$EVTYPE == "RIVER FLOOD", ]$EVTYPE = "FLOOD"
data[data$EVTYPE == "HURRICANE/TYPHOON", ]$EVTYPE = "HURRICANE-TYPHOON"
data[data$EVTYPE == "HEAT", ]$EVTYPE = "EXCESSIVE HEAT"
```

##Results

Here we analyse four types of losses:

1. Fatalities
2. Injuries
3. Property Damages
4. Crop Damages

```{r}

##Fatalities

fatalities <- aggregate(FATALITIES ~ EVTYPE, data = data, sum, na.rm = TRUE)
fatal <- fatalities[fatalities$FATALITIES > 0,]
fatalityorder <- fatal[order(fatal$FATALITIES, decreasing = TRUE),]
head(fatalityorder)


barplot(fatalityorder[1:10, 2], col = rainbow(10), legend.text = fatalityorder[1:10,
    1], ylab = "Fatalities", main = "Top 10 natural events causing most fatalities")

##Injuries

injuries <- aggregate(INJURIES ~ EVTYPE, data = data, sum, na.rm = TRUE)
injury <- injuries[injuries$INJURIES > 0,]
injuryorder <- injury[order(injury$INJURIES, decreasing = TRUE),]
head(injuryorder)

barplot(injuryorder[1:10, 2], col = rainbow(10), legend.text = injuryorder[1:10,
    1], ylab = "Injuries", main = "Top 10 natural events causing most injuries")
```



## Data shows that Tornado is the  major cause of most fatalities and injuries.

```{r}
##Property Damage

data[data$PROPDMGEXP == "K", ]$PROPDMG <- data[data$PROPDMGEXP == "K", ]$PROPDMG * 1000
data[data$PROPDMGEXP == "M", ]$PROPDMG <- data[data$PROPDMGEXP == "M", ]$PROPDMG * 1e+06
data[data$PROPDMGEXP == "m", ]$PROPDMG <- data[data$PROPDMGEXP == "m", ]$PROPDMG * 1e+06
data[data$PROPDMGEXP == "B", ]$PROPDMG <- data[data$PROPDMGEXP == "B", ]$PROPDMG * 1e+09

propertydamages <- aggregate(PROPDMG ~ EVTYPE, data = data, sum, na.rm = TRUE)
damage <- propertydamages[propertydamages$PROPDMG > 0,]
damageorder <- damage[order(damage$PROPDMG, decreasing = TRUE),]
head(damageorder)

barplot(damageorder[1:10, 2], col = rainbow(10), legend.text = damageorder[1:10,
    1], ylab = "Property Damages", main = "Top 10 natural events causing most property damages")

##Crop Damage

data[data$CROPDMGEXP  == "K", ]$CROPDMG <- data[data$CROPDMGEXP  == "K", ]$CROPDMG * 1000
data[data$CROPDMGEXP  == "M", ]$CROPDMG <- data[data$CROPDMGEXP  == "M", ]$CROPDMG * 1e+06
data[data$CROPDMGEXP  == "m", ]$CROPDMG <- data[data$CROPDMGEXP  == "m", ]$CROPDMG * 1e+06
data[data$CROPDMGEXP  == "B", ]$CROPDMG <- data[data$CROPDMGEXP  == "B", ]$CROPDMG * 1e+09

cropdamages <- aggregate(CROPDMG ~ EVTYPE, data = data, sum, na.rm = TRUE)
cropdamage <- cropdamages[cropdamages$CROPDMG > 0,]
cropdamageorder <- cropdamage[order(cropdamage$CROPDMG, decreasing = TRUE),]
head(cropdamageorder)
```

## Data shows that Flood is the  major cause of property damages and Drought is the major reason for crop damages.
