---
title: "ReproducibleResearch - Project1"
author: "Lalitha"
date: "December 5, 2017"
output: html_document
---
## Load the data

```{r}
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
unzip(temp)

data <- read.csv("activity.csv")
```

##What is mean total number of steps taken per day?

Calculate total number of steps taken per day and build a histogram

```{r}
 steps_per_day <- aggregate(steps ~ date, data , sum)
 hist(steps_per_day$steps, main = paste("Total Steps taken per day"), col= "green", xlab = "Number of Steps")
```
##Calculate mean and median of the total number of steps taken per day

```{r}
  rmean <- mean(steps_per_day$steps, na.rm = TRUE)
  rmedian <- median(steps_per_day$steps, na.rm = TRUE)
```

The mean is `r rmean`. The median is `r rmedian`

##What is the average daily pattern?

```{r}
steps_by_interval <- aggregate(steps ~ interval, data, mean, na.rm=TRUE)
plot(steps_by_interval$interval, steps_by_interval$steps, type = "l", xlab = "5-minute Interval", ylab = "Average Number of steps", main = "Average Number of Steps taken per day by Interval")

max_steps_by_Interval <- steps_by_interval[which.max(steps_by_interval$steps),1]
```
The maximum number of steps, on average across all the days in the dataset in a 5-minute interval is `r max_steps_by_Interval`

##Imputing missing values

```{r}

 ##Total number of missing values
    missingvalues <- is.na(data$steps)
    table(missingvalues)
  
##Difference   
imputed_data <- transform(data,steps= ifelse(is.na(data$steps), steps_by_interval$steps[match(data$interval,steps_by_interval$interval)], data$steps))
imputed_data[as.character(imputed_data$date) == "2012-10-01", 1] <- 0

steps_per_day_i <- aggregate(steps ~ date, imputed_data, sum)
hist(steps_per_day_i$steps, main = paste("Total Steps Each Day"), col="red", xlab="Number of Steps")

hist(steps_per_day$steps, main = paste("Total Steps taken per day"), col= "green", xlab = "Number of Steps", add=TRUE)
legend("topright", c("Imputed", "Non-imputed"), col=c("red", "green"), lwd=10)

rmean.i <- mean(steps_per_day_i$steps)
rmedian.i <- median(steps_per_day_i$steps)

```
The imputed data mean is `r rmean.i`.  The imputed data median is `r rmedian.i`. The non-imputed data mean is `r rmean`. The non-imputed data median is `r rmedian`

##Are there differences in activity patterns between weekdays and weekends?
```{r}
##Identify weekdays and weekend
   weekdays <- c("Monday","Tuesday","Wednesday","Thursday","Friday")
imputed_data$dow = as.factor(ifelse(is.element(weekdays(as.Date(imputed_data$date)),weekdays),"weekday","weekend"))
 
steps_per_day_i <-aggregate(steps ~ interval+dow, imputed_data, mean) 

library(lattice)

xyplot(steps_per_day_i$steps ~ steps_per_day_i$interval|steps_per_day_i$dow, main="Average Steps per Day  by Interval",xlab="Interval", ylab="Number of Steps",layout=c(1,2), type="l")

```

